---
title: "Impact Indicators from ENAHO"
subtitle: "RBM 2022 Baselines"
author: "Bruno Carayon, IM Officer Costa Rica"
date: "`r format(Sys.Date(),  '%d %B %Y')`"
output:
  unhcrdown::pptx_slides
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.cap = TRUE)

library(officedown)
library(tidyverse)
library(unhcrthemes)
library(fontawesome)
library(haven)
library(srvyr)

# turn off the automatic use of showtext functionality, so that the dml function can works properly
showtext::showtext_auto(FALSE)

data <- read_sav(here::here("data-raw", "ENAHO 2021 usuarios .sav")) %>%
  as_factor()

```

```{r tidying, include=FALSE}

# Create New Variables : poverty, drinking water, sanitation, hygiene, electricity,clean fuels, access2_all_BS, health insurance, primary_enrollment,secondary_enrollment,secure_tenure,CondAct

data <- data |>
  # create variable poverty (poor for extreme poverty and poverty no extreme o no poor)
  mutate(
    poverty = as.factor(
      if_else(
        np == "Pobreza extrema" | np == "Pobreza no extrema",
        "Poor",
        "No poor"
      )
    )
  ) |>
  mutate(
    drinking_water =
      if_else(V12 == "Pozo" | V12 == "Río, quebrada o naciente" | V12 == "Lluvia u otro", "No", "Yes")
  ) |>
  mutate(sanitation = if_else(V13A == "No tiene", "No", "Yes")) |>
  mutate(hygiene = if_else(V14A == "No", "No", "Yes")) |>
  mutate(electricity = if_else(V15 == "No hay luz eléctrica", "No", "Yes")) |>
  mutate(clean_fuels = if_else(V16 == "Ninguno (no cocina)" | V16 == "Leña o carbón" | V16 == "Otro", "No", "Yes")) |>
  mutate(access2_all_BS = ifelse(
    drinking_water == "Yes" &
      sanitation == "Yes" &
      hygiene == "Yes" &
      electricity == "Yes" &
      clean_fuels == "Yes",
    "Yes",
    "No"
  )) |>
  mutate(health_insurance = if_else(
    A11 == "No asegurado",
    "No",
    "Yes"
  )) |>
  mutate(primary_enrollment = if_else(
    A13 == "Escuela" & (A5 == 6 | A5 == 7 | A5 == 8 |
      A5 == 9 | A5 == 10 | A5 == 11),
    "Yes",
    "No"
  )) |>
  mutate(secondary_enrollment = if_else(
    A13 == "Colegio" & (A5 == 12 | A5 == 13 | A5 == 14 |
      A5 == 15 | A5 == 16 | A5 == 17),
    "Yes",
    "No"
  )) |>
  mutate(secure_tenure = if_else(V2A == "Propia totalmente pagada" | V2A == "Propia pagando a plazos" | V2A == "Alquilada", "Yes", "No")) |>
  mutate(A5 = as.numeric(A5))


svy <- as_survey(
  data,
  strata = c(REGION, ZONA),
  ids = c(UPM, CUESTIONARIO),
  weights = FACTOR
)
```

---

# Inclusión estadística

Tal y como promueven las [Recomendaciones Internacionales sobre Estadísticas de Refugiados (IRRS)](https://egrisstats.org/recommendations/international-recommendations-on-refugee-statistics-irrs/) de 2018,las Oficinas Nacionales de Estadística son las mejores posicionadas para producir datos de alta calidad sobre el desplazamiento forzado.

Los indicadores generados a partir de las estadísticas oficiales son particularmente relevantes no solo para informar el diseño de programas humanitarios a corto plazo, sino también la formulación de políticas y la financiación del desarrollo a largo plazo

El ACNUR dispone de un conjunto de indicadores básicos. En Costa Rica, la Encuesta Nacional de Hogares incluye a las personas desplazadas por la fuerza y el ACNUR puede utilizar las estadísticas oficiales para calcular sus propios indicadores.



---

# Encuesta Nacional de Hogares

La [Encuesta Nacional de Hogares (ENAHO)](https://inec.cr/estadisticas-fuentes/encuestas/encuesta-nacional-hogares) es un programa de recolección de datos cuyo enfoque está asociado al nivel de bienestar de la población, especialmente centrados en la conformación del ingreso de los hogares, su distribución y características de los hogares y la población en situación de pobreza. 

Se contempla además el estudio de la tenencia de vivienda y sus características, el acceso de las personas a la educación y al seguro social, así como la población que trabaja y las condiciones de esos trabajos, entre otros. 

Igualmente, cada año se hacen investigaciones específicas o módulos especiales, dentro de los que se pueden mencionar acceso a programas sociales utilización de servicios de salud, telecomunicaciones, trabajo infantil, migración, seguridad ciudadana, lactancia materna.

---  

# Impact Area 2 - Realizing Basic Rights in Safe Environment




```{r ia2, warning=FALSE, layout='Title and Content', ph=officer::ph_location_type(type="body")}


# 2.1 Proportion of PoC living below the national poverty line. [GCR 2.1.1 and SDG 1.2.1]

ind_21_nic <- svy |>
  filter(LugNac == "En Nicaragua") |>
  group_by(poverty) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(poverty == "Poor")

ind_21_crc <- svy |>
  filter(LugNac == "En mismo cantón de residencia" | LugNac == "En otro cantón de residencia") |>
  group_by(poverty) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(poverty == "Poor")


# 2.2 Proportion of PoCs residing in physically safe and secure settlements with access to basic facilities
## Considering: Access to water(V13A) + Electricity(V15) + Hygiene(V14A) + Clean fuels(V16)  = Person has access to basic services


ind_22_nic <- svy |>
  filter(LugNac == "En Nicaragua") |>
  group_by(access2_all_BS) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(access2_all_BS == "Yes")


ind_22_crc <- svy |>
  filter(LugNac == "En mismo cantón de residencia" | LugNac == "En otro cantón de residencia") |>
  group_by(access2_all_BS) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(access2_all_BS == "Yes")


# 2.3 Proportion of PoC with access to health services
## Considering health insurance as access to health services


ind_23_nic <- svy |>
  filter(LugNac == "En Nicaragua") |>
  group_by(health_insurance) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(health_insurance == "Yes")


ind_23_crc <- svy |>
  filter(LugNac == "En mismo cantón de residencia" | LugNac == "En otro cantón de residencia") |>
  group_by(health_insurance) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(health_insurance == "Yes")


# and now the graph

pop <- c("CostaRicans", "Nicaraguans", "CostaRicans", "Nicaraguans", "CostaRicans", "Nicaraguans")
ind <- c("rbm2.1", "rbm2.1", "rbm2.2", "rbm2.2", "rbm2.3", "rbm2.3")
p <- c(ind_21_crc$p, ind_21_nic$p, ind_22_crc$p, ind_22_nic$p, ind_23_crc$p, ind_23_nic$p)
p_cv <- c(ind_21_crc$p_cv, ind_21_nic$p_cv, ind_22_crc$p_cv, ind_22_nic$p_cv, ind_23_crc$p_cv, ind_23_nic$p_cv)


indicators <- tibble(pop, ind, p, p_cv) %>%
  mutate(ind = as.factor(ind), pop = as.factor(pop))

rbm_indicators <-
  c(
    rbm2.1 = "2.1 Proportion of PoC living below the national poverty line",
    rbm2.2 = "2.2 Proportion of PoCs residing in physically safe and secure settlements with access to basic facilities.",
    rbm2.3 = "2.3 Proportion of PoC with access to health services"
  )

ggplot(
  data = indicators,
  aes(
    x = p,
    y = fct_rev(ind),
    color = pop
  )
) +
  geom_point() +
  geom_errorbar(aes(xmin = p - p_cv, xmax = p + p_cv), width = 0.2) +
  geom_text(
    aes(label = scales::label_percent(.1)(p)),
    vjust = ifelse((indicators$pop == "Nicaraguans"), -1.25, 2)
  ) +
  scale_x_continuous(
    labels = scales::label_percent(),
    limits = c(0, 1),
    breaks = c(0, .5, 1)
  ) +
  scale_color_manual(values = c(
    "#666666",
    "#0072BC"
  )) +
  scale_y_discrete(labels = str_wrap(rev(rbm_indicators), 40)) +
  labs(
    x = "Estimate",
    y = "",
    title = "Impact Area 2:",
    subtitle = " Realizing Rights in Safe Environments",
    caption = glue::glue(
      "Encuesta Nacional de Hogares (ENAHO) 2021\n",
      "# of Nicaraguans = 2,229)"
    )
  ) +
  unhcrthemes::theme_unhcr(font_size = 15) +
  theme(panel.grid.minor.x = element_blank()) -> p


# turn off the automatic use of showtext functionality, so that the dml function can works properly
showtext::showtext_auto(FALSE)
rvg::dml(ggobj = p, fonts = list(serif = "Lato"))
```

--- 

# Impact Area 3 - Empowering Communities and Achieving Gender Equality



```{r ia3, layout='Title and Content', ph=officer::ph_location_type(type="body")}

# Primary education enrollment
## considering individuals from 6 to 11

## Nicaraguans

ind32a_nic <- svy |>
  filter(LugNac == "En Nicaragua") |>
  filter((A5 == 6 | A5 == 7 | A5 == 8 |
    A5 == 9 | A5 == 10 | A5 == 11)) |>
  group_by(primary_enrollment) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(primary_enrollment == "Yes")

## CostaRicans

ind32a_crc <- svy |>
  filter(LugNac == "En mismo cantón de residencia" | LugNac == "En otro cantón de residencia") |>
  filter((A5 == 6 | A5 == 7 | A5 == 8 |
    A5 == 9 | A5 == 10 | A5 == 11)) |>
  group_by(primary_enrollment) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(primary_enrollment == "Yes")

# Secundary education enrollment
## considering individuals from 12 to 17


## Nicaraguans

ind32b_nic <- svy |>
  filter(LugNac == "En Nicaragua") |>
  filter(A5 == 12 | A5 == 13 | A5 == 14 |
    A5 == 15 | A5 == 16 | A5 == 17) |>
  group_by(secondary_enrollment) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(secondary_enrollment == "Yes")

## CostaRicans

ind32b_crc <- svy |>
  filter(LugNac == "En mismo cantón de residencia" | LugNac == "En otro cantón de residencia") |>
  filter(A5 == 12 | A5 == 13 | A5 == 14 |
    A5 == 15 | A5 == 16 | A5 == 17) |>
  group_by(secondary_enrollment) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(secondary_enrollment == "Yes")

# and now the graph

pop <- c("Costa Ricans", "Nicaraguans", "Costa Ricans", "Nicaraguans")
ind <- c("rbm3.2a", "rbm3.2a", "rbm3.2b", "rbm3.2b")
p <- c(ind32a_crc$p, ind32a_nic$p, ind32b_crc$p, ind32b_nic$p)
p_cv <- c(ind32a_crc$p_cv, ind32a_nic$p_cv, ind32b_crc$p_cv, ind32b_nic$p_cv)

indicators <- tibble(pop, ind, p, p_cv) %>%
  mutate(ind = as.factor(ind), pop = as.factor(pop))


rbm_indicators <-
  c(
    rbm3.1a = "3.2a Proportion of PoC enrolled in primary education",
    rbm3.1b = "3.2b Proportion of PoC enrolled in secondary education"
  )

ggplot(
  data = indicators,
  aes(
    x = p,
    y = fct_rev(ind),
    color = pop
  )
) +
  geom_point() +
  geom_errorbar(aes(xmin = p - p_cv, xmax = p + p_cv), width = 0.2) +
  geom_text(
    aes(label = scales::label_percent(.1)(p)),
    vjust = ifelse((indicators$pop == "Nicaraguans"), -1.25, 2)
  ) +
  scale_x_continuous(
    labels = scales::label_percent(),
    limits = c(0, 1),
    breaks = c(0, .5, 1)
  ) +
  scale_color_manual(values = c(
    "#666666",
    "#0072BC"
  )) +
  scale_y_discrete(labels = str_wrap(rev(rbm_indicators), 40)) +
  labs(
    x = "Estimate",
    y = "",
    title = "Impact Area 3:",
    subtitle = "Empowering Communities and Achieving Gender Equality",
    caption = glue::glue(
      "Encuesta Nacional de Hogares (ENAHO) 2021\n",
      "# of Nicaraguans = 2,229)"
    )
  ) +
  unhcrthemes::theme_unhcr(font_size = 15) +
  theme(panel.grid.minor.x = element_blank()) -> p


# turn off the automatic use of showtext functionality, so that the dml function can works properly
showtext::showtext_auto(FALSE)
rvg::dml(ggobj = p, fonts = list(serif = "Lato"))
```


--- 

# Outcome Indicators


```{r outcome, layout='Title and Content', ph=officer::ph_location_type(type="body")}


# 8.2 Proportion of PoC with primary reliance on clean (cooking) fuels and technology [SDG 7.1.2 Tier 1]

# Nicaraguans

ind81_nic <- svy |>
  filter(LugNac == "En Nicaragua") |>
  group_by(clean_fuels) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(clean_fuels == "Yes")


# CostaRicans

ind81_crc <- svy |>
  filter(LugNac == "En mismo cantón de residencia" | LugNac == "En otro cantón de residencia") |>
  group_by(clean_fuels) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(clean_fuels == "Yes")

# 13.3 Proportion of PoC (working age) who are unemployed.
## considering the work force above 15 Unemployed -> CondAct = 2 and Employed -> CondAct == 1

# Nicaraguans

ind133_nic <- svy |>
  filter(LugNac == "En Nicaragua") |>
  filter(A5 > 14) |>
  filter(CondAct == "Ocupado" | CondAct == "Desempleado abierto") |>
  group_by(CondAct) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(CondAct == "Desempleado abierto")



# Costaricans


ind133_crc <- svy |>
  filter(LugNac == "En mismo cantón de residencia" | LugNac == "En otro cantón de residencia") |>
  filter(A5 > 14) |>
  filter(CondAct == "Ocupado" | CondAct == "Desempleado abierto") |>
  group_by(CondAct) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(CondAct == "Desempleado abierto")

# 16.1. Proportion of PoC with secure tenure rights and/or property rights to housing and/or land [revised SDG indicator 1.4.2]
# considering that the property is owned (partially o totally paid) o rented.


# Nicaraguans

ind161_nic <- svy |>
  filter(LugNac == "En Nicaragua") |>
  group_by(secure_tenure) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(secure_tenure == "Yes")

# Costaricains

ind161_crc <- svy |>
  filter(LugNac == "En mismo cantón de residencia" | LugNac == "En otro cantón de residencia") |>
  group_by(secure_tenure) |>
  summarize(p = survey_prop(vartype = "cv")) |>
  filter(secure_tenure == "Yes")


# and now the graph

pop <- c("Costa Ricans", "Nicaraguans", "Costa Ricans", "Nicaraguans", "Costa Ricans", "Nicaraguans")
ind <- c("rbm8.1", "rbm8.1", "rbm13.3", "rbm13.3", "rbm16.1", "rbm16.1")
p <- c(ind81_crc$p, ind81_nic$p, ind133_crc$p, ind133_nic$p, ind161_crc$p, ind161_nic$p)
p_cv <- c(ind81_crc$p_cv, ind81_nic$p_cv, ind133_crc$p_cv, ind133_nic$p_cv, ind161_crc$p_cv, ind161_nic$p_cv)

indicators <- tibble(pop, ind, p, p_cv) %>%
  mutate(ind = as.factor(ind), pop = as.factor(pop))


rbm_indicators <-
  c(
    rbm8.1 = "8.2 Proportion of PoC with primary reliance on clean (cooking) fuels and technology [SDG 7.1.2 Tier 1]",
    rbm16.1 = "16.1. Proportion of PoC with secure tenure rights and/or property rights to housing and/or land [revised SDG indicator 1.4.2]",
    rbm13.3 = "13.3 Proportion of PoC (working age) who are unemployed."
  )

ggplot(
  data = indicators,
  aes(
    x = p,
    y = ind,
    color = pop
  )
) +
  geom_point() +
  geom_errorbar(aes(xmin = p - p_cv, xmax = p + p_cv), width = 0.2) +
  geom_text(
    aes(label = scales::label_percent(.1)(p)),
    vjust = ifelse((indicators$pop == "Nicaraguans"), -1.25, 2)
  ) +
  scale_x_continuous(
    labels = scales::label_percent(),
    limits = c(0, 1),
    breaks = c(0, .5, 1)
  ) +
  scale_color_manual(values = c(
    "#666666",
    "#0072BC"
  )) +
  scale_y_discrete(labels = str_wrap(rev(rbm_indicators), 40)) +
  labs(
    x = "Estimate",
    y = "",
    title = "Outcome Indicators:",
    caption = glue::glue(
      "Encuesta Nacional de Hogares (ENAHO) 2021\n",
      "# of Nicaraguans = 2,229)"
    )
  ) +
  unhcrthemes::theme_unhcr(font_size = 15) +
  theme(panel.grid.minor.x = element_blank()) -> p


# turn off the automatic use of showtext functionality, so that the dml function can works properly
showtext::showtext_auto(FALSE)
rvg::dml(ggobj = p, fonts = list(serif = "Lato"))
```



---

# Thank you!


---
